Naturalness and Long-lived Beyond the Standard Model Particles

### 2017-07-27 tex/draft.tex
Thoroughly edit Naturalness section

\usepackage[draft]{fixme}
\fxsetup{layout=footnote}
\usepackage[left]{lineno}
\usepackage{amsmath,amsthm,amssymb,mathtools,graphicx,caption,subcaption,mathrsfs,cancel,multirow,enumitem,setspace,sectsty,cite,hyperref}
\usepackage{fullpage}

@@

\newpage
\tableofcontents
\newpage
\doublespacing
\linenumbers
\pagenumbering{arabic}

\section{The Standard Model}
    The standard model of particle physics (SM) describes all known particles and their non-gravitational interactions and is experimentally verified to astonishing \fxnote{weak word} accuracy. The SM consists of 12 spin-$\frac{1}{2}$ particles, the fermions, that make up all observed matter; 12 spin-1 particles, the gauge bosons, that communicate the electromagnetic, weak, and strong forces; and 1 fundamental scalar, the Higgs boson, which will be described in detail below.

    All SM particles transform according to the gauged symmetry group $\SUthreecSUtwoLUoneY$, where $\SUthreec$ is associated with color charge, $\SUtwoL$ is associated with weak isospin, and $\UoneY$ is associated with weak hypercharge. The gauge bosons are associated with the generators of these groups, thus there are 8 gluons associated with $\SUthreec$, three W bosons associated with $\SUtwoL$, and one B boson associated with $\UoneY$. The physical gauge bosons, which are linear combinations of the Ws and B, are the $\Wp$, $\Wm$, $\Z$, and photon ($\gamma$). The fermions are catagorized by which representation of each group they fit into. For example, $\SUthreec$ distinguishes the quarks (which live in the fundamental representation) from the leptons (which are singlets), and $\SUtwoL$ distinguishes the left-handed fermions (doublets) from the right-handed fermions (singlets). Finally, weak hypercharge further distignuishes SM particles according to the Gell-Mann--Nishijima formula, $Y=Q-I_3$, where Q is electric charge and $I_3$ is the third component of weak isospin.

@@

    SM parameters, such as particle masses\fxnote{what else is corrected. How?}, are corrected by interactions with virtual particles through loop diagrams. Figure\fxnote{make and reference diagrams}, for example, shows the largest SM corrections to the Higgs and electron masses. Because the Higgs is a fundamental scalar, its corrections depend quadratically on the cutoff scale, $\Lambda$, which is the largest energy scale at which the SM is valid. The fermions and gauge bosons, on the other hand, only depend logarithmically\fxnote{check. be able to work out example} on $\Lambda$. The next section explores the concept of naturalness, a major motivating principle for beyond the standard model (BSM) physics that is directly tied to the relationship between the bare Higgs mass and the terms that correct it.

@@

\section{Naturalness and the Higgs mass}
    \fxnote{cite guidice naturally and t'hooft naturalness}
    \fxnote{Haven't specifically said 'hierarchy problem. Is that bad?}
    \fxnote{Never really talked about the effective theory concept or dimensional analysis}
    The naturalness criterion can be stated in many ways, a few of which are reproduced here: (1) All dimensionless parameters should be of order one, (2) An effective theory should not be sensitive to small variations in the parameters of higher-energy theory, and (3) any dimensionless parameter much smaller than one must be protected by a custodial symmetry. When applied to the Higgs mass, naturalness leads to what's known as the `hierarchy problem' or unacceptable levels of `fine tuning'. These issues are unpacked below.

    Experimentally, the mass of the Higgs boson is 125 GeV, a value that must be achieved by a linear combination of the bare mass and loop diagram corrections. The scale of the loop corrections is proportional to the square of the cutoff scale of the SM, which is the highest energy at which the SM is valid. If, hypothetically, the SM is valid all the way up to the Planck scale (where quantum gravity becomes important), then Higgs mass corrections would be order($MP^2$) $\approx 10^{32}$ GeV.

    \fxnote{Point is that corrections come in with different signs and individual terms are huge. Really weird that they would cancel out so perfectly. Maybe add feynman diagrams to illustrate. Use Jaxodraw probably} Order($10^{32}$) corrections imply that the bare Higgs mass must be negative and also order($10^{32}$) to reproduce the experimetnally known value. In fact, the bare mass and corrections must match to one part in $\approx 10^{30}$. While it's entirely possible that these two numbers just so happen to match so perfectly, many \fxnote{weak} feel that such miraculous fine tuning must be a hint that some deeper physical mechanism is at work. That is, either some previously unknown symmetry is protecting the Higgs mass or the cutoff scale is order(TeV).

    \fxnote{Confusing, says Andrew. Maybe add math}  Before digging into possible solutions to this problem\fxnote{specify}, one other question must be addressed: why doesn't this same argument apply to all SM particles? The answer comes from naturalness statement (3) above. As a fundamental scalar, the Higgs boson lacks the chiral or gauge symmetries enjoyed by the fermions and vector bosons, respectively. These two symmetries cause the symmetry of the SM to be enhanced when the fermion and vector boson masses are set to 0. \fxnote{work out fermion case?} Setting the higgs mass to 0 in no way increases the symmetry of the theory, so we have no reason to expect it to be small.

    \fxnote{Clarify that not technically a problem. Nature might just be that way. Anthropic principle almost like an attempt to restore naturalness to a universe that appears unnatural, says Andrew} There are three main approaches to resolving this problem\fxnote{specify}: (1) argue that it was never really a problem in the first place, (2) reduce the SM cutoff scale, and (3) introduce a new symmetry to protect the Higgs mass. Arguments against naturalness often involve the anthropic principle, which claims that it should come as no surprise that humans live in a universe whose fundamental parameters are tuned to produce carbon-based life. Alternatively, some theories, such as large extra dimensions, lower the SM cutoff by proposing that gravity is spread across more than 3 spatial dimensions, which lowers the Planck mass itself. Finally, the dominant BSM theory that aims to solve the naturalness problem by introducing a new symmetry is supersymmetry, which is explained below. \fxnote{didn't mention composite higgs}

    \fxnote{Optional paragraph about non-Higgs unnaturalness:    There are other unnatural quantities that people worry about in physics. I know of two off the top of my head: (1) thetaQCD in the strong CP problem (2) the cosmological consant. The strong CP problem comes from the fact that QCD technically allows CP violation even though such violation has never been observed. The degree of violation is parameterized as thetaQCD, which recasts the problem into the naturalness context of 'why is thetaQCD so small?'. The cosmological constant is experimentally $\approx$60 to $\approx$120 orders of magnitude smaller than the predicted value \fxnote{predicted how?}. While these issues deserve serious attention, they will not be considered further here}

>>>>

\usepackage[final]{fixme}
\fxsetup{layout=footnote}
\usepackage[left]{lineno}
\usepackage{amsmath,amsthm,amssymb,mathtools,graphicx,caption,subcaption,mathrsfs,cancel,multirow,enumitem,setspace,sectsty,cite,hyperref,siunitx}
\usepackage{fullpage}

@@

\newpage
\tableofcontents
\newpage
\doublespacing
%\linenumbers
\pagenumbering{arabic}

\section{The Standard Model}
    The standard model of particle physics (SM) describes all known particles and their non-gravitational interactions and is experimentally verified to astonishing \fxnote{weak word} accuracy. The SM consists of \num{12} spin-$\frac{1}{2}$ particles, the fermions, that make up all observed matter; \num{12} spin-1 particles, the gauge bosons, that communicate the electromagnetic, weak, and strong forces; and one fundamental scalar, the Higgs boson, which will be described in detail below.

    All SM particles transform according to the gauged symmetry group $\SUthreecSUtwoLUoneY$, where $\SUthreec$ is associated with color charge, $\SUtwoL$ is associated with weak isospin, and $\UoneY$ is associated with weak hypercharge. The gauge bosons are associated with the generators of these groups, thus there are \num{8} gluons associated with $\SUthreec$, three W bosons associated with $\SUtwoL$, and one B boson associated with $\UoneY$. The physical gauge bosons, which are linear combinations of the Ws and B, are the $\Wp$, $\Wm$, $\Z$, and photon ($\gamma$). The fermions are catagorized by which representation of each group they fit into. For example, $\SUthreec$ distinguishes the quarks (which live in the fundamental representation) from the leptons (which are singlets), and $\SUtwoL$ distinguishes the left-handed fermions (doublets) from the right-handed fermions (singlets). Finally, weak hypercharge further distignuishes SM particles according to the Gell-Mann--Nishijima formula, $Y=Q-I_3$, where Q is electric charge and $I_3$ is the third component of weak isospin.

@@

    SM parameters, such as particle masses\fxnote{what else is corrected. How?}, are corrected by interactions with virtual particles through loop diagrams. Figure\fxnote{make and reference diagrams}, for example, shows the potential corrections to the Higgs and electron masses. Because the Higgs is a fundamental scalar, its corrections depend quadratically on the cutoff scale, $\Lambda$, which is the largest energy scale for which the SM is valid. The fermions and gauge bosons, on the other hand, only depend logarithmically\fxnote{check. be able to work out example} on $\Lambda$. The next section explores the concept of naturalness, a major motivating principle for beyond the standard model (BSM) physics that is directly tied to the relationship between the bare Higgs mass and the terms that correct it.

@@

\section{Naturalness and the Higgs mass}
    \fxnote{cite guidice naturally and t'hooft naturalness}
    \fxnote{Never mentioned dimensional analysis}

    The naturalness criterion can be formulated in many ways, a few of which are reproduced here: (1) All dimensionless parameters should be of order one, (2) An effective theory should not be sensitive to small variations in the parameters of the higher-energy theory, and (3) any dimensionless parameter much smaller than unity must be protected by a custodial symmetry. Physicists implicitly, and perhaps unwittingly, apply naturalness in the form of (1) any time they use dimensional analysis to estimate a quantity, while (2) merely restates the idea that a predictive effective theory requires low-energy phenomena to be decoupled from the details of the high-energy theory\fxnote{This is my weakest point because I don't understand it as well as I'd like to. It's also wordy and maybe repetative}. This discussion focuses on statement (3), which is in some ways a more nuanced version of (1) that explains cases where dimensional analysis fails.

    All massive particles correct the Higgs mass with terms proportional to $\Lambda^2$. Experimentally, the Higgs mass is \SI{125}{GeV}\fxnote{cite CMS, ATLAS}, a value that must be achieved by the sum of the bare mass and all corrections. Although corrections differ in sign, the precise cancellation of many extremely large terms to produce a value as small as the Higgs mass is extremely unlikely. In fact, the fine-tuning necessary to produce the Higgs mass is on the order of one part in \num{e30} if the SM is valid up to the Plank scale ($\mPlanck\approx$ \SI{e16}{GeV}), where quantum gravity becomes important. While it's possible that the bare mass and corrections just so happen to almost perfectly cancel, such miraculous fine tuning could also be a hint that some deeper physical mechanism is at work. That is, either some previously unknown symmetry is protecting the Higgs mass or the cutoff scale is order(TeV).

    As a fundamental scalar, the Higgs is doubly cursed. As explained in \fxnote{reference SM section}, $\mHiggs$ is quadratically sensitive to $\Lambda$, but that's not all: the Higgs field also lacks the chiral and gauge symmetries enjoyed by the fermions and gauge bosons. These symmetries act as the custodial symmetries mentioned in naturalness statement (3) in that they protect the fermion and gauge boson masses from large corrections. Because the SM chiral and gauge symmetries only apply to the massless theory, the SM symmetry group would be enlarged if fermions and gauge bosons were massless. That is, if the fermions and gauge bosons were massless at tree level, the chiral and gauge symmetries would cause their masses to be uncorrected to all orders in perturbation theory. As massive particles, however, the chiral and gauge symmetries instead guarantee that any corrections to their tree-level masses are proportional to their tree-level masses, preventing any large fine tuning between correction terms.

    \fxnote{Clarify that not technically a problem. Nature might just be that way. Anthropic principle almost like an attempt to restore naturalness to a universe that appears unnatural, says Andrew} There are three main approaches to the apparently unnatural Higgs mass\fxnote{specify}: (1) argue that it was never really a problem in the first place, (2) reduce the SM cutoff scale, and (3) introduce a new symmetry to protect the Higgs mass. Arguments against naturalness often involve the anthropic principle, which claims that it should come as no surprise that humans live in a universe whose fundamental parameters are tuned to produce carbon-based life. Alternatively, some theories, such as large extra dimensions, lower the SM cutoff by proposing that gravity is spread across more than 3 spatial dimensions, which lowers the Planck mass itself. Finally, the dominant BSM theory that aims to solve the naturalness problem by introducing a new symmetry is supersymmetry, which is explained below.
